Think deep for reef resilience

### 2025-09-10 paper/paper_draft.qmd
Add spatial dhw sentence

Reefs in different temporal clusters largely do not differ in mean modelled DHW levels
experienced and exhibit overall similar DHW time series. While these reefs experience
similar magnitude DHWs based on SST, the proportion of the reefs' coral population impacted
by heat stress events depends on reef depth. Reefs at deeper depths are likely to be
clustered into High cover clusters. Reefs at shallow depths experience no, or few, years
with coral cover exceeding levels likely required to maintain positive carbonate budgets.
Movement of shallow reefs towards negative carbonate budgets in the future may contribute to
phase shifts in reef dominant CaCO~3~ production. Shifting from coral driven to encrusting
algae driven production, or to bioerosion dominated systems with rapid erosion of physical
reef frameworks [@dudgeon2010; @perry2008]. Reefs at increased depths may be able to
maintain coral driven CaCO~3~ production and accumulation into the future for longer. Our
results suggest that within shallow-water reef ecosystems (non-mesophotic, depth \< 30m),
depth contributes greatly to reef resilience and the ability of the reef to maintain
positive carbonate budgets under future conditions. Spatially our results indicate that
positive carbonate budget maintenance, under ACCESS-ESM1-5, and high depth levels were
concentrated in the Far Northern, offshore Townsville/Whitsunday, and Mackay/Capricorn
management areas (@sfig-carb-budget-map; @sfig-depth-map). Additionally similar spatial
patterns were observed for reefs clustered into the high cover cluster category across all
GCMs (@sfig-gcm-cluster-map). We do not interpret these spatial findings as accurate
predictions of future resilience due to high uncertainty in input data, particularly depth
and connectivity. While similar findings have been demonstrated in the Pacific [@smith2014;
@baird2018] and Indo-Pacific [@bridge2014] regions, depth refugia analyses have often
focussed on the potential of mesophotic reefs (\> 30m deep) as thermal refugia
[@bongaerts2010; @holstein2015; @semmler2017]. In contrast, Venegas et al. [-@venegas2019]
found no association between depth and the alleviation of heat stress across a range of
shallow reefs in the Pacific.

>>>>

Reefs in different temporal clusters largely do not differ in mean modelled DHW levels
experienced and exhibit overall similar DHW time series. While these reefs experience
similar magnitude DHWs based on SST, the proportion of the reefs' coral population impacted
by heat stress events depends on reef depth. Reefs at deeper depths are likely to be
clustered into High cover clusters. Reefs at shallow depths experience no, or few, years
with coral cover exceeding levels likely required to maintain positive carbonate budgets.
Movement of shallow reefs towards negative carbonate budgets in the future may contribute to
phase shifts in reef dominant CaCO~3~ production. Shifting from coral driven to encrusting
algae driven production, or to bioerosion dominated systems with rapid erosion of physical
reef frameworks [@dudgeon2010; @perry2008]. Reefs at increased depths may be able to
maintain coral driven CaCO~3~ production and accumulation into the future for longer. Our
results suggest that within shallow-water reef ecosystems (non-mesophotic, depth \< 30m),
depth contributes greatly to reef resilience and the ability of the reef to maintain
positive carbonate budgets under future conditions. Spatially our results indicate that
positive carbonate budget maintenance, under ACCESS-ESM1-5, and high depth levels were
concentrated in the Far Northern, offshore Townsville/Whitsunday, and Mackay/Capricorn
management areas (@sfig-carb-budget-map; @sfig-depth-map). Resilient reefs in the Far
Northern management area were exposed to lower mean DHW levels than other reefs under
ACCESS-ESM1-5 (@sfig-mean-dhw-map). Similar spatial patterns were observed across GCMs for
reefs clustered into the high cover cluster category (@sfig-gcm-cluster-map). We do not
interpret these spatial findings as accurate predictions of future resilience due to high
uncertainty in input data, particularly depth and connectivity. While similar findings have
been demonstrated in the Pacific [@smith2014; @baird2018] and Indo-Pacific [@bridge2014]
regions, depth refugia analyses have often focussed on the potential of mesophotic reefs (\>
30m deep) as thermal refugia [@bongaerts2010; @holstein2015; @semmler2017]. In contrast,
Venegas et al. [-@venegas2019] found no association between depth and the alleviation of
heat stress across a range of shallow reefs in the Pacific.
